Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-02-03 paper/rei.tex
Updates from Overleaf

\section{Background}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}. We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis. 
We also test its robustness by taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.
[Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

@@

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

>>>>

%
%
%
%
%

\section{Background}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}. We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.

%
%
\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis. 
We also test its robustness by taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.
[Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

@@

 \begin{figure}
          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{Reionization PCs for $\zmax = 30$.} 
          \label{fig:xe}
\end{figure}

%
%
%
%

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}
